Appendix S1

### 2020-08-07 analysis/paper/appendix/appendix_s1.Rmd
made some formatting changes and minor edits for submission to Ecosphere.

---
title: "Appendix S1"
subtitle: "Estimating Carrying Capacity for Chinook Parr using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::pdf_document2:
      fig_caption: yes
      fig_height: 8.5
      fig_width: 6.5
      toc: no
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks2.lua
      - --lua-filter=../../templates/pagebreak.lua
      includes:
        in_header: ../../templates/header_appendixS.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks.lua
      - --lua-filter=../../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      fig_height: 8.5
      fig_width: 6.5
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks.lua
      - --lua-filter=../../templates/pagebreak.lua
bibliography:
- ../references.bib
- ../packages.bib
csl: "../../templates/ecological-applications.csl" # Insert path for the bib-style
---

@@

```{r packages}
# setwd('analysis/paper')

@@

# options for table formating with "kable()"
options(knitr.kable.NA = '-')
# options(knitr.table.format = "pandoc")

@@

```{r mine-stats}
#-----------------------------------------------------------------
# generate MINE statistics
#-----------------------------------------------------------------
mine_res = tibble(species = unique(fish_hab$Species),
                  fh_data = list(fish_hab)) %>%
  mutate(metrics = map(fh_data,
                       .f = function(x) {
                         poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           pull(ShortName)
                       })) %>%
  mutate(mine_res = map2(fh_data,
                         metrics,
                         .f = function(x, y) {
                           if(sum(x$fish_dens == 0) == 0) {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           } else {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens + 0.005)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           }
                         })) %>%
  select(species, mine_res) %>%
  unnest(cols = mine_res) %>%
  left_join(poss_hab_mets,
            by = c('Metric' = 'ShortName')) %>%
  select(species, MetricCategory, Metric, Name, everything()) %>%
  # put the metric names in descending order by MIC
  mutate_at(vars(Metric, Name),
            list(~ fct_reorder(., .x = MIC))) %>%
  arrange(species, MetricCategory, desc(MIC)) %>%
      # filter out some metrics with too many NAs or 0s
      # filter((perc_NA < 0.5 & non_0 > 100)) %>%
      # filter out metrics with very low variance
      # filter(var < 0.1) %>%
      # filter(obsCV < 0.1)
      # filter out area and volume metrics
      filter(!grepl('Area$', Metric),
             !grepl('Vol$', Metric),
             !Metric %in% c('Lgth_Wet', 
                            'Lgth_BfChnl',
                            'Lgth_WetChnl',
                            'Area_Wet', 
                            'Area_Bf', 
                            'WetVol', 
                            'BfVol')) %>%
  mutate(selected = if_else(Metric %in% sel_hab_mets$Metric,
                            T, F))

@@

hab_corr = tibble(hab_data = list(hab_data)) %>%
  mutate(metrics = map(hab_data,
                       .f = function(x) {
                         mets = poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           pull(ShortName)
                         x %>%
                           select(one_of(mets)) %>%
                           gather(Metric, value) %>%
                           group_by(Metric) %>%
                           summarise(n_tot = n(),
                                     n_NA = sum(is.na(value)),
                                     non_0 = sum(value != 0, na.rm = T)) %>%
                           mutate(perc_NA = n_NA / n_tot,
                                  perc_non_0 = non_0 / n_tot) %>%
                           filter(perc_NA < 0.5,
                                  non_0 > 100) %>%
                           filter(!grepl('Area$', Metric),
                                  !grepl('Vol$', Metric),
                                  !Metric %in% c('Lgth_Wet', 
                                                 'Lgth_BfChnl',
                                                 'Lgth_WetChnl',
                                                 'Area_Wet', 
                                                 'Area_Bf', 
                                                 'WetVol', 
                                                 'BfVol')) %>%
                           pull(Metric)
                       }),
         cor_mat = map2(hab_data,
                        metrics,
                        .f = function(x, y) {
                          x %>%
                            select(one_of(y)) %>%
                            cor(use = 'pairwise',
                                method = 'pearson')
                                # method = 'spearman')
                        }),
         corr_mat = map(cor_mat,
                        .f = as_cordf),
         catg_cor = map2(hab_data,
                         metrics,
                         .f = function(x, y) {
                           catg_list = try(poss_hab_mets %>%
                                             filter(ShortName %in% y) %>%
                                             split(list(.$MetricCategory)) %>%
                                             map(.f = function(z) {
                                               if(nrow(z) == 1) {
                                                 return(NULL)
                                               } else {
                                                 x %>%
                                                   select(one_of(z$ShortName)) %>%
                                                   cor(use = 'pairwise',
                                                       method = 'pearson')
                                                       # method = 'spearman')
                                               }
                                               }))
                           return(catg_list)
                                             }))

@@

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figure \@ref(fig:mine-facet-fig)). The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Figure \@ref(fig:corr-fig)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by rehabilitation actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

@@

mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        # format = 'html',
        # format = 'latex',
        booktabs = T,
        longtable = T,
        caption = 'MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T,
                latex_options = c('repeat_header',
                                  'scale_down',
                                  'HOLD_position')) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T) %>%
  column_spec(2, width = "2in") %>%
  column_spec(3, width = "1in") %>%
  column_spec(c(4:6), width = "0.5in")

@@

```{r mine-all-fig, fig.cap = 'Barplot of MIC statistics, colored by habitat category.'}
ggplot(data = mine_res,
       aes(x = Name,
           y = MIC,
           fill = MetricCategory)) +
  geom_col() +
  coord_flip() +
  # scale_fill_viridis_d() +
  scale_fill_brewer(palette = 'Set3',
                    guide = guide_legend(nrow = 2)) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 6)) +
  labs(fill = 'Category')
```

@@

```{r corr-fig, eval = T, results='hide', out.width = "90%", fig.show = "hold", fig.cap = "Correlation plots of metrics, faceted by habitat category. Larger circles correspond to larger correlation coefficients. The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

```{r high-cor-mets, fig.cap = 'Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5.'}
high_cor_mets = hab_data %>%
  select(one_of(sel_hab_mets$Metric)) %>%
  correlate(use = 'pairwise',
            method = 'pearson') %>%
            # method = 'spearman') %>%
  stretch(na.rm = T,
          remove.dups = T) %>%
  arrange(desc(abs(r))) %>%
  filter(abs(r) > 0.5) %>%
  select(x, y) %>%
  pivot_longer(cols = c(x, y)) %>%
  pull(value) %>%
  unique()

@@

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

>>>>

---
title: "Appendix S1"
subtitle: "Estimating Carrying Capacity for Juvenile Salmon using Quantile Random Forest Models"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [biomark]
      correspondence: true
  - Michael W. Ackerman:
      email: Mike.Ackerman@merck.com
      institute: [biomark]
      correspondence: false
  - Richard A. Carmichael:
      email: Richard.Carmichael@merck.com
      institute: [biomark]
      correspondence: false
  - Sarah L. Hoffmann:
      email: Sarah.Hoffmann@merck.com
      institute: [biomark]
      correspondence: false
  - Chris Beasley:
      email: Chris.Beasley@merck.com
      institute: [biomark]
      correspondence: false
institute:
  - biomark: Biomark, Inc. 705 South 8th St., Boise, Idaho, 83702, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      toc: no
      always_allow_html: true
      number_sections: FALSE
      fig_height: 7
      fig_width: 6
      reference_docx: "../../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks.lua
      - --lua-filter=../../templates/pagebreak.lua
    bookdown::pdf_document2:
      fig_caption: yes
      fig_height: 8.5
      fig_width: 6.5
      toc: no
      number_sections: FALSE
      geometry: margin=1in
      latex_engine: "xelatex"
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks2.lua
      - --lua-filter=../../templates/pagebreak.lua
      includes:
        in_header: ../../templates/header_appendixS.tex
    bookdown::html_document2:
      fig_caption: yes
      fig_height: 8.5
      fig_width: 6.5
      number_sections: FALSE
      pandoc_args:
      - --lua-filter=../../templates/scholarly-metadata.lua
      - --lua-filter=../../templates/author-info-blocks.lua
      - --lua-filter=../../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- ../references.bib
- ../packages.bib
csl: "../../templates/ecological-applications.csl" # Insert path for the bib-style
---

@@

```{r packages}
# setwd('analysis/paper/appendix')

@@

# options for table formating with "kable()"
options(knitr.kable.NA = '-')

# what kind of document is being created?
doc.type <- knitr::opts_knit$get('rmarkdown.pandoc.to')
# when knitting to Word, use this
if(doc.type == 'docx') {
  options(knitr.table.format = "pandoc")
}

# cat(doc.type)
```

@@

```{r mine-stats}
#-----------------------------------------------------------------
# generate MINE statistics
#-----------------------------------------------------------------
mine_res = tibble(species = unique(fish_hab$Species),
                  fh_data = list(fish_hab)) %>%
  mutate(metrics = map(fh_data,
                       .f = function(x) {
                         poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           filter(ShortName != "ChnlUnitTotal_Ct") %>%
                           pull(ShortName)
                       })) %>%
  mutate(mine_res = map2(fh_data,
                         metrics,
                         .f = function(x, y) {
                           if(sum(x$fish_dens == 0) == 0) {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           } else {
                             try(x %>%
                                   mutate(fish_dens = log(fish_dens + 0.005)) %>%
                                   estimate_MIC(covars = y,
                                                response = 'fish_dens'))
                           }
                         })) %>%
  select(species, mine_res) %>%
  unnest(cols = mine_res) %>%
  left_join(poss_hab_mets,
            by = c('Metric' = 'ShortName')) %>%
  select(species, MetricCategory, Metric, Name, everything()) %>%
  # put the metric names in descending order by MIC
  mutate_at(vars(Metric, Name),
            list(~ fct_reorder(., .x = MIC))) %>%
  arrange(species, MetricCategory, desc(MIC)) %>%
      # filter out some metrics with too many NAs or 0s
      # filter((perc_NA < 0.5 & non_0 > 100)) %>%
      # filter out metrics with very low variance
      # filter(var < 0.1) %>%
      # filter(obsCV < 0.1)
      # filter out area and volume metrics
      filter(!grepl('Area$', Metric),
             !grepl('Vol$', Metric),
             !Metric %in% c('Lgth_Wet', 
                            'Lgth_BfChnl',
                            'Lgth_WetChnl',
                            'Area_Wet', 
                            'Area_Bf', 
                            'WetVol', 
                            'BfVol')) %>%
  mutate(selected = if_else(Metric %in% sel_hab_mets$Metric,
                            T, F))

@@

hab_corr = tibble(hab_data = list(hab_data)) %>%
  mutate(metrics = map(hab_data,
                       .f = function(x) {
                         mets = poss_hab_mets %>%
                           filter(ShortName %in% names(x)) %>%
                           filter(ShortName != "ChnlUnitTotal_Ct") %>%
                           pull(ShortName)
                         x %>%
                           select(one_of(mets)) %>%
                           gather(Metric, value) %>%
                           group_by(Metric) %>%
                           summarise(n_tot = n(),
                                     n_NA = sum(is.na(value)),
                                     non_0 = sum(value != 0, na.rm = T)) %>%
                           mutate(perc_NA = n_NA / n_tot,
                                  perc_non_0 = non_0 / n_tot) %>%
                           filter(perc_NA < 0.5,
                                  non_0 > 100) %>%
                           filter(!grepl('Area$', Metric),
                                  !grepl('Vol$', Metric),
                                  !Metric %in% c('Lgth_Wet', 
                                                 'Lgth_BfChnl',
                                                 'Lgth_WetChnl',
                                                 'Area_Wet', 
                                                 'Area_Bf', 
                                                 'WetVol', 
                                                 'BfVol')) %>%
                           pull(Metric)
                       }),
         cor_mat = map2(hab_data,
                        metrics,
                        .f = function(x, y) {
                          x %>%
                            select(one_of(y)) %>%
                            cor(use = 'pairwise',
                                method = 'pearson')
                                # method = 'spearman')
                        }),
         corr_mat = map(cor_mat,
                        .f = as_cordf),
         catg_cor = map2(hab_data,
                         metrics,
                         .f = function(x, y) {
                           catg_list = try(poss_hab_mets %>%
                                             filter(ShortName %in% y) %>%
                                             split(list(.$MetricCategory)) %>%
                                             map(.f = function(z) {
                                               if(nrow(z) == 1) {
                                                 return(NULL)
                                               } else {
                                                 x %>%
                                                   select(one_of(z$ShortName)) %>%
                                                   cor(use = 'pairwise',
                                                       method = 'pearson')
                                                       # method = 'spearman')
                                               }
                                               }))
                           return(catg_list)
                                             }))

# create data.frame with correlation coefficients and habitat categories
corrr_df = hab_corr$catg_cor[[1]] %>%
  map(.f = as_cordf) %>%
  map(.f = stretch,
      remove.dups = T,
      na.rm = T) %>%
  map_df(.id = 'Category',
         .f = identity) %>%
  left_join(poss_hab_mets %>%
              select(x = ShortName,
                     `Metric 1` = Name)) %>%
  left_join(poss_hab_mets %>%
              select(y = ShortName,
                     `Metric 2` = Name)) %>%
  select(Category,
         `Metric 1`,
         `Metric 2`,
         r) %>%
  mutate_at(vars(Category),
            list(str_replace_all),
            pattern = "([:lower:])([:upper:])",
            replacement = "\\1 \\2") %>%
  arrange(Category,
          desc(abs(r)))

@@

Within each category, metrics were ranked according to their MIC value (Table \@ref(tab:mic-tab) and Figures \@ref(fig:mine-facet-fig) and \@ref(fig:mine-all-fig)). The MIC value of each measured habitat characteristic and parr density was used to inform decisions on which habitat covariates to include in the QRF parr capacity model. We selected one or two variables amongst those with the highest MIC scores within each category, attempting to avoid covariates that were too highly correlated (Table \@ref(tab:corr-tab)), while focusing on covariates we thought could influence fish behavior. For example, cumulative drainage area, mean annual flow and observed discharge are all highly correlated, but fish really only experience the observed discharge, so we chose to include that metric in our QRF model. We also tried to include covariates that can be directly influenced by rehabilitation actions or have been shown to impact salmonid juvenile density. Finally, we attempted to avoid metrics with too many missing values, or too many zero values, in the data set, as well as metrics that may have too much observer error [@Rosgen2018].

# Results

We chose `r nrow(sel_hab_mets)` metrics, highlighted in bold in Table \@ref(tab:mic-tab). Their correlations with each other are displayed in Figure \@ref(fig:selected-corr). Those with a high correlation coefficient ($\geq$ 0.5 or $\leq$ -0.5) are plotted against each other in Figure \@ref(fig:high-cor-mets) to show the variety in values even for pairs of metrics with relatively high correlations.

@@

if (doc.type != "docx") {
mine_tab_df %>%
  select(-selected) %>%
  kable(digits = 3,
        booktabs = T,
        longtable = T,
        caption = 'MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold.') %>%
  kable_styling(fixed_thead = T,
                latex_options = c('repeat_header',
                                  'scale_down',
                                  'HOLD_position')) %>%
  row_spec(row = which(mine_tab_df$selected),
           bold = T) %>%
  column_spec(2, width = "2in") %>%
  column_spec(3, width = "1in") %>%
  column_spec(c(4:6), width = "0.5in")
}

if(doc.type == 'docx') {
  library(pander)
  mine_tab_df %>%
    select(-selected) %>%
    pander(digits = 3,
           caption = '(\\#tab:mic-tab)MIC statistic for top metrics within each habitat category, sorted by category and MIC value. The percent of records for which each habitat metric measurement was missing or zero is also shown. Metrics selected for the QRF model are in bold.',
           emphasize.strong.rows = which(mine_tab_df$selected),
           split.tables = Inf)
}

```{r corr-tab}
new_catg_rows = corrr_df %>%
  mutate(lead_catg = lead(Category),
         new_catg = if_else(Category != lead_catg, 
                            T, F)) %>%
  pull(new_catg) %>%
  which()

if (doc.type == "latex") {
  corrr_df %>%
    kable(digits = 2,
          booktabs = T,
          longtable = T,
          format = "latex",
          caption = "Pearson correlation coefficient between each variable within a habitat category.") %>%
    kable_styling(fixed_thead = T,
                  latex_options = c('repeat_header',
                                    'scale_down',
                                    'HOLD_position')) %>%
    row_spec(new_catg_rows,
             hline_after = T) %>%
    column_spec(1, width = "1in") %>%
    column_spec(c(2, 3), width = "2in")
}

if (doc.type == "html") {
  corrr_df %>%
    kable(digits = 2,
          booktabs = T,
          # longtable = T,
          caption = "Pearson correlation coefficient between each variable within a habitat category.") %>%
    kable_styling(fixed_thead = T,
                  latex_options = c('repeat_header',
                                    'scale_down',
                                    'HOLD_position')) %>%
    collapse_rows(columns = 1,
                  latex_hline = "major",
                  valign = "top")
}

if(doc.type == 'docx') {
  library(pander)

  corrr_df %>%
    pander(digits = 2,
           caption = "(\\#tab:corr-tab)Pearson correlation coefficient between each variable within a habitat category.",
           split.tables = Inf)

}

@@

```{r mine-all-fig, fig.cap = 'Barplot of MIC statistics, colored by habitat category.'}
mine_res %>%
    mutate_at(vars(MetricCategory),
            list(str_replace_all),
            pattern = "([:lower:])([:upper:])",
            replacement = "\\1 \\2") %>%

ggplot(aes(x = Name,
           y = MIC,
           fill = MetricCategory)) +
  geom_col() +
  coord_flip() +
  # scale_fill_viridis_d() +
  scale_fill_brewer(palette = 'Set3',
                    guide = guide_legend(ncol = 3)) +
  theme(legend.position = 'bottom',
        axis.text = element_text(size = 6)) +
  labs(fill = 'Category')
```

@@

```{r corr-fig, eval = F, results='hide', out.width = "90%", fig.show = "hold", fig.cap = "Correlation plots of metrics, divided by habitat category. Larger circles correspond to larger correlation coefficients. The colors correspond to the correlation coefficient as well, and range from blue (-1) to white (0) to red (1)."}
cor_plot_df = tibble(Category = names(hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]),
                     cor_mat = hab_corr$catg_cor[[1]][!sapply(hab_corr$catg_cor[[1]], is.null)]) %>%
  mutate(cor_plot = map2(cor_mat,
                         Category,
                         .f = function(x, y) {
                           ggcorrplot::ggcorrplot(x,
                                                  method = 'circle',
                                                  hc.order = T,
                                                  type = 'lower',
                                                  title = y,
                                                  colors = c('blue4', 'white', 'red3'))
                         }))

@@

```{r high-cor-mets, fig.cap = 'Pairs plot of habitat metrics used in QRF model with a correlation coefficient greater than 0.5 or less than -0.5.'}
high_cor_mets = hab_data %>%
  select(one_of(sel_hab_mets$Metric)) %>%
  correlate(use = 'pairwise',
            method = 'pearson') %>%
            # method = 'spearman') %>%
  stretch(na.rm = T,
          remove.dups = T) %>%
  arrange(desc(abs(r))) %>%
  filter(abs(r) > 0.5) %>%
  select(x, y) %>%
  pivot_longer(cols = c(x, y)) %>%
  pull(value) %>%
  unique()

@@
